---
output:
  html_document: default
  pdf_document: default
---
White Wine Quality EDA by Adrián Vera Ros
========================================================
## Introduction
In this EDA we aim to analyze the white variants of the Portuguese "Vinho Verde"
wine with data from the book "Modeling wine preferences by data mining from 
physicochemical properties." by P. Cortez, A. Cerdeira, F. Almeida, T. Matos 
and J. Reis.

Doing a preliminary research about the "Vinho Verde" denomination, we discover
that it's an Origin Denomination (OD) and not a grape variety. "Vinho verde" 
translates as "Young wine", or wines that are relased 3-6 months after harvest. 

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide', packages}
install.packages("ggplot2", dependencies = T, 
                 repos = "http://cran.us.r-project.org")
install.packages("knitr", dependencies = T, 
                 repos = "http://cran.us.r-project.org")
install.packages("dplyr", dependencies = T, 
                 repos = "http://cran.us.r-project.org")
install.packages("PerformanceAnalytics", dependencies = T, 
                 repos = "http://cran.us.r-project.org")
install.packages("memisc", dependencies = T, 
                 repos = "http://cran.us.r-project.org")

```

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide', libraries}
library(ggplot2)
library(PerformanceAnalytics)
library(scales)
library(gridExtra)
library(memisc)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
whitewine <- read.csv('wineQualityWhites.csv')
```


## Univariate Plots Section
The dataset is comprised by 4898 observations from 13 different variables, where
we have the wines ordered by a numeric key, different objective physicochemical 
characteristics of each wine and the median of at least 3 subjective sensory
evaluations made by wine experts. 

More information about the nature of the variables can be found here: https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt


```{r echo=FALSE, message=FALSE, warning=FALSE, Dataset_analyis}
str(whitewine)


summary(whitewine)
p1 <- qplot(x = fixed.acidity, data = whitewine)
p2 <- qplot(x = volatile.acidity, data = whitewine)
p3 <- qplot(x = citric.acid, data = whitewine)
p4 <- qplot(x = alcohol, data = whitewine)
p5 <- qplot(x = density, data = whitewine)
p6 <- qplot(x = residual.sugar, data = whitewine)
p7 <- qplot(x = chlorides, data = whitewine)
p8 <- qplot(x = free.sulfur.dioxide, data = whitewine)
p9 <- qplot(x = total.sulfur.dioxide, data = whitewine)
p10 <- qplot(x = sulphates, data = whitewine)
p11 <- qplot(x = pH, data = whitewine)
p12 <- qplot(x = quality, data = whitewine)



grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12,  ncol = 3)

```

At first glance we can detect outliers in our dataset, both in the summary 
(with the max value far the average and the 3rd quartile) and in the graphs. 

Attending to the distributions, most variables have a normal distribution not 
centered in the graph due to these outliers. However, in the case of residual 
sugar and alcohol, this situation doesn't happen and we find a left skewed 
distribution. 

### Outliers elimination and plot transformation

```{r echo=FALSE, data_cleaning}
wwt <- subset(whitewine, total.sulfur.dioxide <= 250 & sulphates <= 0.8 
              & free.sulfur.dioxide <= 80 & residual.sugar <= 20 & density <= 1 
              & citric.acid <= 0.7 & volatile.acidity <= 0.6 
              & chlorides <= 0.075 & fixed.acidity <= 10)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, sugar_trans}
ggplot(aes(x=residual.sugar), data = wwt)+
  geom_histogram() +
  scale_x_continuous(breaks = seq(0,65,5)) +
  scale_y_continuous(breaks = seq(0,500,100), trans = "log10")

ggplot(aes(x=alcohol), data = wwt)+
  geom_histogram(binwidth = 0.25) +
  scale_x_continuous(breaks = seq(0,20,0.5)) +
  scale_y_continuous(breaks = seq(0,500,100), trans = "log10")
```

After transforming the variables we notice how sugar still mantains an skewed
distribution and how alcohol has spikes in the distribution, probably due to 
some figures having been rounded.


### Quality
```{r echo=FALSE, Quality}
wwt.qualitymean <- mean(wwt$quality)
wwt$quality.cat <- cut(as.numeric(wwt$quality)
                         ,c(2, 4, 6, 10)
                         ,labels = c("Low", "Medium", "High"))

ggplot(aes(x=quality), data = wwt)+
  geom_vline(xintercept = wwt.qualitymean, color = "red") +
  scale_x_continuous(breaks = 0:10)+
  geom_bar()

ggplot(aes(x=quality.cat), data = wwt)+
  geom_bar()



summary(wwt$quality)
table(wwt$quality)
    
```

The wine score are integers, so there are no scores like 6,5. 
It ranges from 3 to 9, with 6 being the median with 2012 observations and a 
average score of 5,9. 
Of the 4407 observations we have after eliminating the outliers, 3.09% have a 
score between 3 and 4 (Low), 74.11% have a score between 5 and 6 (Medium) and 
22,8% have a score of 7 or more (High). However, only 3,77% have a score of 8 
or more. 

```{r echo=FALSE, message=FALSE, warning=FALSE, trimmed_dataset_analysis}
str(wwt)
summary(wwt)
p1t <- qplot(x = fixed.acidity, data = wwt)
p2t <- qplot(x = volatile.acidity, data = wwt)
p3t <- qplot(x = citric.acid, data = wwt)
p5t <- qplot(x = density, data = wwt)
p6t <- qplot(x = chlorides, data = wwt)
p7t <- qplot(x = free.sulfur.dioxide, data = wwt)
p8t <- qplot(x = total.sulfur.dioxide, data = wwt)
p9t <- qplot(x = sulphates, data = wwt)
p10t <- qplot(x = pH, data = wwt)



grid.arrange(p1t, p2t, p3t, p5t, p6t, p7t, p8t, p9t, p10t, ncol = 2)

```



## Univariate Analysis
Most of the wines are between 9 and 11% alcohol and most of our dataset is 
composed by average wines while only a few score of 8 and more. 

According to information coming with the dataset, most wines have a pH level 
between 3 and 4, which concurs with our wine's pH normal distribution. 

According to EU regulation 753/2002, wine's like the ones we have in our
dataset can be considered medium or medium dry depending on if the sweetness
gets balanced with acidity.  

Finally, the dataset states that the lower the percentage of alcohol and sugar, 
the closest the density of the wine is to water. During the process of 
fermentation, the sugar is converted to alcohol, resulting on lower levels of 
density. The opposite to a wine with high density (or watery) would be a wine 
with gravity (with higher sugar/alcohol content). 


While the full alcohol range in our dataset is between 8 and 14 degrees, most of
the wines fall between 9 and 11 degrees of alcohol. Since "Vinho Verde" is an
OD and not only a type of grape, we are willing to accept the asumption that
different grapes and fermentations processes can result in different alcohol 
and residual sugar contents.

Sulphates additives help preventing wine oxidation and preserving the wine's 
freshness through higher sulfur dioxide (SO2) gas levels. However, too much SO2
will have implications on the wine taste and potentially, on the consumer's 
health. 

Small quantities of citric acid adds freshness to the flavor. Too much citric
acid can create an unpleasant vinegar-like taste.

Salt is commonly used in the culinary world as a flavor enhacer, which might
affect the wine final flavor and rating. 

All of the above variables are different measures that can affect wine flavor, 
so they will be the main characteristics for the analysis in the following
stages. 

So what's next? 
First we are going to observe the correlation between variables to make sure
that we are selecting the most relevant for the wine rating analysis. 

Then, we will transform quality to a factor variable to be able to perform a 
better bivariate analysis and observe the relationship between characteristics
and score and between the different characteristics. 






## Bivariate Plots Section

```{r echo=FALSE, correlation1}

cor(x=wwt[,2:12], y=wwt$quality)
```

Here we can see the main correlation scores to quality. We now select the most
important ones and create a correlation chart with them.

```{r echo=FALSE, main_Correlation}
main <- c("quality","total.sulfur.dioxide", "residual.sugar", 
          "volatile.acidity", "alcohol", "chlorides", "density")
whitefinal <- wwt[main]

chart.Correlation(whitefinal, histogram = TRUE, pch = 1)
```

In the correlation chart we can observe the strong relationship between density
and alcohol/sugar, as stated in the dataset documentation. Now is time to first
analyze alcohol to wine quality score (as the main factor) and then to other
relevant characteristics. 

```{r echo=FALSE, Bivariate_plots_quality}
whitefinal$quality <- factor(whitefinal$quality, 
                             levels=c(0,1,2,3,4,5,6,7,8,9,10))

ggplot(data=whitefinal, aes(x=quality, y=alcohol)) +
    geom_jitter(alpha=0.2) +
    geom_boxplot(alpha = .5,color = 'blue') +
    stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```





```{r echo=FALSE, Bivariate_plots_quality2}
b1 <- ggplot(data=whitefinal, aes(x=as.numeric(quality), y=alcohol)) +
    geom_jitter(alpha=0.2) +
    geom_smooth(method='lm', aes(group = 1))+
    xlab("Wine Quality") +
    ylab("SO2")

b2 <- ggplot(data=whitefinal, aes(x=as.numeric(quality), 
                                  y=total.sulfur.dioxide)) +
    geom_jitter(alpha=0.2) +
    geom_smooth(method='lm', aes(group = 1))+
    xlab("Wine Quality") +
    ylab("SO2")

b3 <- ggplot(data=whitefinal, aes(x=as.numeric(quality), y=residual.sugar)) +
    geom_jitter(alpha=0.2) +
    geom_smooth(method='lm', aes(group = 1))+
    xlab("Wine Quality") +
    ylab("Sugar")

b4 <- ggplot(data=whitefinal, aes(x=as.numeric(quality), y=volatile.acidity)) +
    geom_jitter(alpha=0.2) +
    geom_smooth(method='lm', aes(group = 1))+
    xlab("Wine Quality") +
    ylab("Acidity")

b5 <- ggplot(data=whitefinal, aes(x=as.numeric(quality), y=chlorides)) +
    geom_jitter(alpha=0.2) +
    geom_smooth(method='lm', aes(group = 1))+
    xlab("Wine Quality") +
    ylab("Salt")

b6 <- ggplot(data=whitefinal, aes(x=as.numeric(quality), y=density)) +
    geom_jitter(alpha=0.2) +
    geom_smooth(method='lm', aes(group = 1))+
    xlab("Wine Quality") +
    ylab("Density")

grid.arrange(b1, b2, b3, b4, b5, b6, ncol = 2)

```



## Bivariate Analysis
Surprisingly, alcohol seems to be the main driver on the quality score, with a 
positive correlation of 0.44. Salt, SO2 and sugar, on the other hand, have soft 
negative correlations with the score. At the sime time, these three variables 
have a moderate negative correlations with the alcohol % content and
positive with density.

We can corroborate that density is opposite to alcohol and sugar in our chart,
but we can also see that wines with high density (or low gravity) get worse 
scores. 

Finally, the acidity measure seems to be mild, with no middle or strong
relationship whatsoever.


In the next stage (multivariate analysis) we are going to investigate the 
relationship between the variables on a more complex level and try to build a 
predictive model to rate white wines. 






## Multivariate Plots Section

First we create a new variable, quality.cat, to classify the wines into three
categories depending on the score. 

Low quality: Scores between 3 and 4
Medium quality: Scores between 5 and 6
High quality: scores between 7 and 9

```{r echo=FALSE, cat}

whitefinal$quality.cat <- cut(as.numeric(whitefinal$quality)
                         ,c(2, 5, 7, 10)
                         ,labels = c("Low", "Medium", "High"))

summary(whitefinal$quality.cat)

```

Then we analyze the different relationships between quality, alcohol and other
variables

```{r echo=FALSE, quaalcvar}
ggplot(data = whitefinal,
       aes(y = residual.sugar, x = alcohol,
           color = quality.cat)) +
  geom_jitter(alpha=.1, size=3) +
  geom_smooth(method = "lm", se = FALSE,size=1)

ggplot(data = whitefinal,
       aes(y = total.sulfur.dioxide, x = alcohol,
           color = quality.cat)) +
  geom_jitter(alpha=.1, size=3) +
  geom_smooth(method = "lm", se = FALSE,size=1)

ggplot(data = whitefinal,
       aes(y = volatile.acidity, x = alcohol,
           color = quality.cat)) +
  geom_jitter(alpha=.1, size=3) +
  geom_smooth(method = "lm", se = FALSE,size=1)

```

In the graph volatile acidity/ alcohol / quality graph above it can be seen how 
only for High quality wines there is a positive relationship between alcohol %
content and acidity. 

Now we will build our model:

```{r echo=FALSE, LM}
m1<-lm(as.numeric(quality) ~ alcohol, data = whitefinal)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + total.sulfur.dioxide)
m4 <- update(m3, ~ . + residual.sugar)
m5 <- update(m4, ~ . + chlorides)
mtable(m1,m2,m3,m4,m5)

```




## Multivariate Analysis

From our plots we get that there seems to be a trade-off between alcohol and 
sugar, sulfates and acidity; where either the wines are high in alcohol content
or in sugar/sulfates/acidity. 

It can be seen as well as wines with higher quality usually have a higher amount
of both parameters (alcohol + sugar, alchol + sulfates or alcohol + volatile 
acidity), as there seems to be a higher presence of quality wines on the upper
right quadrants. 

The lack of data on the right upper quadrant in the relationship between 
sugar and alcohol makes sense since during the fermentation, the sugar in the 
grapes becomes alcohol. High quality wines seem to have a sugar content higher 
than medium and low quality wines.  

Wines with lower sulfates also seem to have lower quality scores than the 
average. 

However, we can find that the combination of alcohol and acidity seems to hit 
the spot regarding the palate of our experts. According to Peynaud, a French 
oenologist, a dry wine can taste sweet if the alcohol level is elevated. 
This seems to be the make or break regarding our wines, as the high score wine
show a tendence of having a balance betweeen both acidity and presence. 

In the graph it can be seen as the upper right quadrant gets more importance on 
visualization; confirming that the combination of high levels of alcohol and 
acidity result in better wines.

Finally, after creating the linear model we can see disappointing results: the 
model built can only explain 25% of the score. This might be due not only to the 
underlying variable connections or the unbalanced dataset where most of the 
wines where average, but also due to the complexity of the wine rating. With no 
clear criteria on how to separate great wine from average or bad, it is not 
possible to create a predictive model. 

------

## Final Plots and Summary

We are now going to summarize the dataset exploration through three graphs and
draw conclusions from our previous work

### Plot One

```{r echo=FALSE, Plot_One}
ggplot(aes(x=quality.cat, fill = as.numeric(quality.cat)), data = wwt)+
  geom_bar()+
  theme(legend.position="none")+
  ggtitle("Distribution of wine quality")+
  xlab("Wine Quality") +
  ylab("Samples")
```

### Description One
Since we aim to analyze the quality of "Vinho Verde" white wine, the first step
would be to know how the quality is distributed on our dataset. Here we can see
how the wine scores are distributed using our category 

Of the 4407 observations we have after eliminating the outliers, 3.09% have a 
score of 3 to 4, 74.11% have a score between 5 and 6 and 22,8% have a 
score of 7 or more. However, only 3,77% have a score of 8 or more. This makes
our dataset to be composed mostly by wines of medium quality, which might be
a difficulty in order to find the characteristics that separates medium quality
wines from premium wines. 


### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data=whitefinal, aes(x=quality, y=alcohol)) +
    geom_jitter(alpha=0.15) +
    geom_boxplot(alpha = .5,color = 'blue') +
    stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
               xlab("Wine Quality") +
               ylab("Alcohol %")+
               ggtitle("Alcohol % per wine quality score")
  
```

### Description Two
In this second plot we can see the interaction between alcohol and wine quality,
after finding that alcohol was the main driver of wine quality score, with a 
correlation of 0.44. 

It can be seen how the alcohol content is higher in wines with a higher quality
score. However, in the case of wines of high quality, we see how the average 
score is lower than the median, and even falls ouside of the box. This makes us
think that we need a higher sample of high quality wines in order to find
a more equal distribution. 


### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = whitefinal,
       aes(y = volatile.acidity, x = alcohol,
           color = quality.cat)) +
  geom_jitter(alpha=.15, size=3) +
  geom_smooth(method = "lm", se = FALSE,size=1)+
  scale_colour_discrete(name="Quality \nCategory")+
          xlab("Alcohol %") +
          ylab("acetic acid (g / dm^3)")+
          ggtitle("Alcohol % and Volatile Acidity per Wine Quality Category")
```

### Description Three
Finally, in this graph it can be seen the relationship between alcohol and 
volatile acidity, other of the variables with higher correlation with wine 
quality score. What makes this graph stand out is that shows more wines with 
high scores in the right upper quadrant than the other multivariate analysis,
signaling a positive relationship between alcohol and acidity in the case of
great wines. 


## Reflection

Regarding this analysis, the main hindrances were the faulty dataset and the
complexity of how the variables are intertwined. Even analyzing the variables 
with higher correlation with the score we have been able to develop only a model
able to explain only a 25% of the score, and it's still not clear to me how
to separate good wine from bad.

Maybe a larger dataset, with more observations of extreme quality scores and 
and the mean score and not the median would be more useful for this kind of 
analysis. In that sense, I find hard to believe that the alcohol is the main 
driver of wine taste and I wish I knew more about wines to be able to create 
relationship between similar variables, like acidity, sulphates and such. Maybe
that way I could do a better job determining what makes a great wine. 

Overall, I enjoyed doing this project but it leaves a sour aftertaste.